Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens

### 2014-01-29 limits/ms.tex
Minor edits.

\begin{abstract}
We use a new non-parametric gravitational modelling tool -- \Glass\ -- to
determine what quality of data (strong lensing, stellar kinematics, and/or
stellar masses) are required to measure the circularly averaged mass profile of
a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass
pixels to model the lens, searching through thousands of models to marginalise
over model uncertainties. Our key findings are as follows: (i) for pure lens
data, multiple sources with wide redshift separation give the strongest
constraints as this breaks the well-known mass-sheet or steepness degeneracy;
(ii) a single quad with time delays also performs well, giving a good recovery
of both the mass profile and its shape; (iii) stellar masses -- for lenses
where the stars dominate the central potential -- can also break the steepness
degeneracy, giving a recovery almost as good as having time delay data or
multiple source redshifts; (iv) stellar kinematics provide a robust measure of
the mass at the half light radius of the stars $r_{1/2}$ that can also break
the steepness degeneracy if $r_{1/2} \neq r_E$ -- the Einstein radius; and (v)
if $r_E \sim r_{1/2}$, then stellar kinematic data can be used to probe the
stellar velocity anisotropy $\beta$ -- an interesting quantity in its own
right. Where information on the mass distribution from lensing and/or other
probes becomes redundant, this opens up the possibility of using strong lensing
to constrain cosmological models. 
\end{abstract}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing.\footnote{These
  quantities are defined more precisely in \S\ref{sec:lensing_basic}.}
First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology.  Typically:
%
\begin{equation}
D_L \approx z \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
%
For multiple images, the sky-projected density must exceed the critical
density
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{D_S}{D_{LS}} \Sigma_L = \frac{D_S}{D_{LS}} \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
%
in some region.  The separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$) is the gravitational radius. If the source is a
quasar or otherwise rapidly variable, a time delay $\Delta t$ in the variability
will be present where:
%
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\item {\bf Stellar kinematic constraints:} This was first suggested by
  \citet{2002MNRAS.337L...6T} as a means to break lensing
  degeneracies. This works because stellar kinematics can provide an
  independent estimate of the Einstein radius, via the virial theorem:
%
\begin{equation}
  \frac{\langle v^2_\mathrm{los}\rangle}{c^2} \approx
  \frac{\theta_E}{6\pi} \frac{D_S}{D_{LS}}
\label{eqn:virial}
\end{equation}
%
where $\langle v^2_\mathrm{los}\rangle$ is the projected stellar
velocity dispersion, and the above relation becomes exact for
isothermal lenses.  The technique has been applied to many lenses
\citep{2006ApJ...649..599K,2008ApJ...682..964B}.  Going further, the
use of two-dimensional kinematics \citep{2011MNRAS.415.2215B} is
especially interesting.

@@

The lens equation:
%
\begin{equation}
\begin{aligned}
    \vec\beta &= \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta) \\
\vec\alpha(\vec\theta) &= \frac{4G}{c^2D_L} \int \Sigma(\vec\theta')
                          \frac{(\vec\theta - \vec\theta')}
                          {\ |\vec\theta - \vec\theta'|^2} \, d^2\vec\theta'
\end{aligned}
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position
$\vec\beta$.  The lens can be thought of as a projected surface
density $\Sigma$ which diverts the path of a photon instantaneously
through the bending angle $\vec\alpha$.  The $D$ factors, as in the
previous section, are angular diameter distances, which depend on the
cosmological density-parameters $\Omega$, the redshifts $z_L,z_S$ of
the lens and the source, and the Hubble parameter $H_0$, thus
%
\begin{equation}
D_{LS} = \frac c{H_0} \frac{1+z_S}{1+z_L} \int_{z_L}^{z_S}
                      \frac{dz}{\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}}
\end{equation}
%
and $D_L \equiv D_{0,L}$, $D_S \equiv D_{0,S}$.  One way to understand the lens
equation is via Fermat's principle. We can think of light as travelling only
along extremum paths where lensed images occur.  Such paths occur at the
extrema of the photon {\it arrival time} $t(\vec\theta)$ that depends on the
geometric path the photon takes and the general relativistic gravitational time
dilation due to a thin lens at redshift $z_L$:
%
\begin{equation}
\begin{aligned}
\frac{ct(\vec\theta)}{(1+z_L)D_{L}}
&= {\textstyle\frac12} |\vec\theta - \vec\beta|^2
   \times \frac{D_{S}}{D_{LS}} \\
&- \frac{4GD_L}{c^2}
   \int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| \, d^2\vec\theta'
\label{full arrival time}
\end{aligned}
\end{equation}
%
We can simplify the above equation by introducing a dimensionless time $\tau$
and density $\kappa$: 
%
\begin{equation}
\tau(\vec\theta) = \frac{ct(\vec\theta)}{(1+z_L)D_{L}} \quad ; \quad \kappa(\vec\theta) \equiv \frac{\Sigma(\vec\theta)}{\Sigma_L}
\end{equation}
%
and hence rewrite \eqnref{full arrival time} as:
%
\begin{equation}
\tau(\vec\theta) = {\textstyle\frac12} |\vec\theta - \vec\beta|^2
                   \times \frac{D_{S}}{D_{LS}}
                 - \frac1\pi \int \kappa (\vec\theta')
                   \ln|\vec\theta - \vec\theta'| d^2\vec\theta'
\label{arrival time}
\end{equation}
%
The scaled arrival time $\tau$ is like a solid angle. It is of order the area
(in steradians) of the full lensing system. The expression $|\vec\theta -
\vec\beta|^2$ is of order the image-separation squared, and the other terms are
of similar size.  For this reason, is convenient to measure $\tau$ in
arcsec$^{2}$.

>>>>

\begin{abstract}
We use a new non-parametric gravitational modelling tool -- \Glass\ -- to
determine what quality of data (strong lensing, stellar kinematics, and/or
stellar masses) are required to measure the circularly averaged mass profile of
a lens and its shape. \Glass\ uses an under-constrained adaptive grid of mass
pixels to model the lens, searching through thousands of models to marginalise
over model uncertainties. Our key findings are as follows: (i) for pure lens
data, multiple sources with wide redshift separation give the strongest
constraints as this breaks the well-known mass-sheet or steepness degeneracy;
(ii) a single quad with time delays also performs well, giving a good recovery
of both the mass profile and its shape; (iii) stellar masses -- for lenses
where the stars dominate the central potential -- can also break the steepness
degeneracy, giving a recovery almost as good as having time delay data or
multiple source redshifts; (iv) stellar kinematics provide a robust measure of
the mass at the half light radius of the stars $r_{1/2}$ that can also break
the steepness degeneracy if the Einstein radius $r_E \neq r_{1/2}$; and (v)
if $r_E \sim r_{1/2}$, then stellar kinematic data can be used to probe the
stellar velocity anisotropy $\beta$ -- an interesting quantity in its own
right. Where information on the mass distribution from lensing and/or other
probes becomes redundant, this opens up the possibility of using strong lensing
to constrain cosmological models. 
\end{abstract}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing.\footnote{These
  quantities are defined more precisely in \S\ref{sec:lensing_basic}.}
First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology.  Typically:
%
\begin{equation}
D_L \approx z \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
%
For multiple images, the sky-projected density must exceed the critical
density
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
%
in some region.  The separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$) is the gravitational radius. If the source is a
quasar or otherwise rapidly variable, a time delay $\Delta t$ in the variability
will be present where:
%
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\item {\bf Stellar kinematic constraints:} This was first suggested by
  \citet{2002MNRAS.337L...6T} as a means to break lensing
  degeneracies. This works because stellar kinematics can provide an
  independent estimate of the Einstein radius, via the virial theorem:
%
\begin{equation}
  \frac{\langle v^2_\mathrm{los}\rangle}{c^2} \approx
  \frac{\theta_E}{6\pi} \frac{D_S}{D_{LS}}
\label{eqn:virial}
\end{equation}
%
where $\langle v^2_\mathrm{los}\rangle$ is the line of sight stellar
velocity dispersion, and the above relation becomes exact for
isothermal lenses.  The technique has been applied to many lenses
\citep{2006ApJ...649..599K,2008ApJ...682..964B}.  Going further, the
use of two-dimensional kinematics \citep{2011MNRAS.415.2215B} is
especially interesting.

@@

The lens equation:
%
\begin{equation}
\begin{aligned}
    \vec\beta &= \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta) \\
\vec\alpha(\vec\theta) &= \frac{4G}{c^2D_L} \int \Sigma(\vec\theta')
                          \frac{(\vec\theta - \vec\theta')}
                          {\ |\vec\theta - \vec\theta'|^2} \, d^2\vec\theta'
\end{aligned}
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position
$\vec\beta$.  The lens can be thought of as a projected surface
density $\Sigma$ which diverts the path of a photon instantaneously
through the bending angle $\vec\alpha$.  The $D$ factors, as in the
previous section, are angular diameter distances, which depend on the
cosmological density-parameters $\Omega$, the redshifts $z_L,z_S$ of
the lens and the source, and the Hubble parameter $H_0$, thus
%
\begin{equation}
D_{LS} = \frac c{H_0} \frac{1+z_S}{1+z_L} \int_{z_L}^{z_S}
                      \frac{dz}{\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}}
\end{equation}
%
and $D_L \equiv D_{0,L}$, $D_S \equiv D_{0,S}$.  One way to understand the lens
equation is via Fermat's principle. We can think of light as travelling only
along extremum paths where lensed images occur.  Such paths occur at the
extrema of the photon {\it arrival time} $t(\vec\theta)$ that depends on the
geometric path the photon takes and the general relativistic gravitational time
dilation due to a thin lens at redshift $z_L$:
%
\begin{equation}
\begin{aligned}
\frac{ct(\vec\theta)}{(1+z_L)D_{L}}
&= {\textstyle\frac12} |\vec\theta - \vec\beta|^2
   \times \frac{D_{S}}{D_{LS}} \\
&- \frac{4GD_L}{c^2}
   \int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| \, d^2\vec\theta'
\label{full arrival time}
\end{aligned}
\end{equation}
%
We can simplify the above equation by introducing a dimensionless time $\tau$
and density $\kappa$: 
%
\begin{equation}
\tau(\vec\theta) = \frac{ct(\vec\theta)}{(1+z_L)D_{L}} \quad ; \quad 
\kappa(\vec\theta) \equiv \frac{\Sigma(\vec\theta)}{\Sigma_\mathrm{crit}}
\end{equation}
%
and hence rewrite \eqnref{full arrival time} as:
%
\begin{equation}
\tau(\vec\theta) = {\textstyle\frac12} |\vec\theta - \vec\beta|^2
                   \times \frac{D_{S}}{D_{LS}}
                 - \frac1\pi \int \kappa (\vec\theta')
                   \ln|\vec\theta - \vec\theta'| d^2\vec\theta'
\label{arrival time}
\end{equation}
%
The scaled arrival time $\tau$ is like a solid angle. It is of order the area
(in steradians) of the full lensing system. The expression $|\vec\theta -
\vec\beta|^2$ is of order the image-separation squared, and the other terms are
of similar size.  For this reason, is convenient to measure $\tau$ in
arcsec$^{2}$.
